\pdftitle

### 2017-05-14 maxNt.tex
add discrepancy section

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-14T12:06:05+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varEta}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varEta'}
\newcommand*{\yHb}{\varEta''}
%Lagr multipliers
\newcommand*{\yL}{\varLambda}
\newcommand*{\yl}{\lambda}
\newcommand*{\yk}{\kappa}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

Our uncertainty about the state of the population is completely expressed by the
probability distribution
\begin{equation}
  \label{eq:joint_plaus}
  \p(X_1=R_1, X_2=R_2, \dotsc, X_N=R_N \| \yH) \quad\text{or}\quad
\p(\yX =\yR \| \yH), \quad \yR \in \set{0,1}^N,
\end{equation}
where $\yH$ denotes our state of knowledge (the evidence and assumptions
backing our probability assignments). Our uncertainty about the state of
the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \p(x_1=r_1, x_2=r_2, \dotsc, x_n=r_n \| \yH) \quad\text{or}\quad
\p(\yx =\yr \| \yH), \quad \yr \in \set{0,1}^n.
\end{equation}

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \binom{n}{n\yrs}
  \sum_{N \yRf =0}^{N} \binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}\,
  \p(\yXf=\yRf \| \yH).
\end{gather}
The combinatorial proof is the same as for the theory of drawing without
replacement, leading to the hypergeometric distribution, and can be found
in many texts
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
The symmetry of our initial ignorance should intuitively also apply to the
sample. This is indeed the case: the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

Formula~\eqref{eq:subpop_average}, connecting our initial probabilities for
the population and sample averages, has a very important consequence: \emph{our
expectations for all averages of the sample and the full population are the
same}:
\begin{multline}
  \label{eq:expe_products}
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
\\[2\jot]
{}=\sum_{N \yRf=0}^N %m!\, 
\binom{N-m}{N \yRf-m}
\binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
\equiv
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N \yRf=0}^N %m!\, 
\binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
\end{multline}

\iffalse
\begin{align}
  \expe{x_{i_1} \dotsm x_{i_m} \| \yH}
&=
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH},
\\
\intertext{and has an explicit expression in terms of $Q$:}
  \begin{split}
\expe{x_{i_1} \dotsm x_{i_m} \| \yH}
&=
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N A=0}^N %m!\, 
\binom{N A}{m}\, Q(A)
\\
&\equiv
%\frac{(N-m)!}{N!}
\sum_{N A=0}^N %m!\, 
\binom{N-m}{N A-m}
\binom{N}{N A}^{-1} Q(A).
\end{split}
\end{align}
\end{subequations}
\fi
***

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-15T00:15:08+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varEta'}
\newcommand*{\yHb}{\varEta''}
%Lagr multipliers
\newcommand*{\yL}{\varLambda}
\newcommand*{\yl}{\lambda}
\newcommand*{\yk}{\kappa}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

Our uncertainty about the state of the population is completely expressed by the
probability distribution
\begin{equation}
  \label{eq:joint_plaus}
  \p(X_1=R_1, X_2=R_2, \dotsc, X_N=R_N \| \yH) \quad\text{or}\quad
\p(\yX =\yR \| \yH), \quad \yR \in \set{0,1}^N,
\end{equation}
where $\yH$ denotes our initial state of knowledge (the evidence and assumptions
backing our probability assignments). Our uncertainty about the state of
the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \p(x_1=r_1, x_2=r_2, \dotsc, x_n=r_n \| \yH) \quad\text{or}\quad
\p(\yx =\yr \| \yH), \quad \yr \in \set{0,1}^n.
\end{equation}

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}.
\end{gather}
The latter conditional probability is a hypergeometric distribution,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

The symmetry present in the probabilities above is not a physical property
of the neuronal population. It only expresses the symmetry of our initial
ignorance about the population, and does not imply any sort of physical
similarity between the neurons. Subsequent observations may in fact break
this symmetry. The initial symmetry should intuitively also apply to the
sample. This is indeed the case: the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

Our initial knowledge (or ignorance) $\yH$ has two very important
properties derivable from the formulae above. First, \emph{our initial
  expectations for all averages of the sample and the full population are
  the same}:
\begin{equation}
  \label{eq:expe_products}
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
%\\[2\jot]
%{}
=
\binom{n}{m}^{-1}
%\frac{(n-m)!}{n!}
\sum_{n \yrs=0}^n %m!\, 
\binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
% \sum_{N \yRf=0}^N %m!\, 
% \binom{N-m}{N \yRf-m}
% \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
\equiv
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N \yRf=0}^N %m!\, 
\binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
\end{equation}
Second, upon observation of a sample average, say $\yxxs=a$, the updated
expectations for such average in the population \emph{and in any new
  sample} will usually be shifted towards the observed value, as follows
from Bayes's theorem and the formulae above.

These two properties express the fact that the sample is
\emph{representative} of the population. Our initial knowledge $\yH$ thus
assumes that we can make predictions about the population by observing a
sample. Contrast this with an assumption that assigns an independent
probability to each neuron: in this case, the observation of one sample
does not change our initial probability about another; we can't learn from
observation.

\subsection{Specifying the initial probability}
\label{sec:specific_initial_probability}

Formulae \eqref{eq:joint_plaus_N_homog}--\eqref{eq:expe_products} are
constraints on our initial probability assignment, but do not determine it
numerically yet. The probability $\p(\yXf=\yRf \| \yH)$ for the population
average needs to be numerically specified, and by \eqref{eq:subpop_average}
it will determine that of the sample average, $\p(\yxs=\yrs \| \yH)$. If we
numerically specify the latter, the former is not completely specified,
because \eqn~\eqref{eq:subpop_average} linearly constrains $N+1$ unknowns
by only $n+1$ equations in this case.

We may want to specify the probability by enforcing several of its sample
expectations to have specific values, for example $\expe{\yxs}=a$,
$\expe{\yxxs}=b$. But this is still and underdetermined linear problem
because several distributions can have the same desired expectations, as
clear from \eqns~\eqref{eq:expe_products}.

It is at this point that the \me\ method enters the scene. It selects one
distribution, purported to be \enquote{maximally noncommittal}, among those
that have the desired expectations. But we are in a quandary:
formulae~\eqref{eq:expe_products} allow us to apply the method to find the
probability of the population $\p(\yXf=\yRf \| \yH)$, or of the sample
$\p(\yxs=\yrs \| \yH)$; the two applications, however, are
\emph{inequivalent}. They lead to numerically different
$\p(\yxs=\yrs \| \yH)$; see \fig~\ref{fig:diff_maxent_pop_sample}. Which
should we choose?

\begin{figure}[!b]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample.pdf}%
\caption{Example of discrepancy in the probability $\p(\yxs = \yrs \| \yH)$
  constructed by \me\ at the population level followed by sample
  marginalization (blue triangles), and at the sample level (red circles).
  The constraints are the same in both applications:
  $\expe{\yxs \|\yH}=0.45$, $\expe{\yxxs \|\yH}=0.35$, and $N=5\,000$,
  $n=200$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
\iffalse
\begin{align}
  \expe{x_{i_1} \dotsm x_{i_m} \| \yH}
&=
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH},
\\
\intertext{and has an explicit expression in terms of $Q$:}
  \begin{split}
\expe{x_{i_1} \dotsm x_{i_m} \| \yH}
&=
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N A=0}^N %m!\, 
\binom{N A}{m}\, Q(A)
\\
&\equiv
%\frac{(N-m)!}{N!}
\sum_{N A=0}^N %m!\, 
\binom{N-m}{N A-m}
\binom{N}{N A}^{-1} Q(A).
\end{split}
\end{align}
\end{subequations}
\fi
***
